---
title: "CFPB Program Submission"
author: "Author: Bernice Boursiquot"
date: "9/27/2022"
output: 
  word_document:
    toc: true
  
---
# Introduction - R Packages and Data Cleaning

This program is located at this private repo https://github.com/boursiquot/CFPB. Permission will be granted upon request.

For this coding example, the following R libraries are needed:
1) dplyr
2) httr
3) janitor
4) jsonlite
5) knitr
6) rlang
7) scales

```{r Setup - Loading in required libraries,  include = FALSE }

knitr::opts_chunk$set(echo = FALSE, warning = FALSE)

set.seed(1)

library(httr)
library(jsonlite)
library(knitr)
library(dplyr)
library(janitor)
library(rlang)

```

 
```{r Pulling in User-Defined Functions}

stateCrosstabFxn = function(var){


}


```

```{r Pulling in Data, include = FALSE}

stateSite = "https://ffiec.cfpb.gov/v2/data-browser-api/view/csv?states=VA,DC,MD&years=2018&actions_taken=5,6"

data = read.csv(stateSite)

```

The data for this exercise come from CFPB's site - `r stateSite`.

```{r Exploring the Data, include = FALSE}

head(data)

colnames(data)

#Counts by state####
data %>%
  group_by(state_code) %>%
    count()



#Counts by applicant race####
data %>%
  group_by(applicant_race.5) %>%
    count()

data %>%
  group_by(applicant_race_observed) %>%
    count()

data %>%
  count(applicant_race.1) 

data %>%
  group_by(applicant_ethnicity.1) %>%
    count()

```

# Analysis
For this exercise, I am interested in looking at the actions taken on a loan; there are 2 outcomes in the dataset - 5 (File closed for incompleteness) and 6 (Purchased loan). 

```{r }

#Counts by action taken####
kable(data %>%
  group_by(action_taken) %>%
    count() %>%
      mutate(N = scales::comma(n)) %>%
        select(-n), col.names = c("Action Taken", "N"), align = "cc")

#Notes: 5 = File closed for incompleteness; 6 = Purchased loan


```


```{r}

data2 =
  data %>%
    mutate(race = (if_else(applicant_race.1 == 1 
                     & (applicant_race.2 %in% c(2,21:27, 3, 4, 41:44, 5) |
                        applicant_race.3 %in% c(2,21:27, 3, 4, 41:44, 5) |
                        applicant_race.4 %in% c(2,21:27, 3, 4, 41:44, 5) |
                        applicant_race.5 %in% c(2,21:27, 3, 4, 41:44, 5))|
                              
                                      applicant_race.1 %in% c(2, 21:27) 
                                    & (applicant_race.2 %in% c(1, 3, 4, 41:44, 5)|
                                         applicant_race.3 %in% c(1, 3, 4, 41:44, 5)|
                                         applicant_race.4 %in% c(1, 3, 4, 41:44, 5)|
                                         applicant_race.5 %in% c(1, 3, 4, 41:44, 5))|
                                      
                                      applicant_race.1 %in% c(3) 
                                    & (applicant_race.2 %in% c(1, 2, 21:27, 4, 41:44, 5)|
                                         applicant_race.3 %in% c(1, 2, 21:27, 4, 41:44, 5)|
                                         applicant_race.4 %in% c(1, 2, 21:27, 4, 41:44, 5)|
                                         applicant_race.5 %in% c(1, 2, 21:27, 4, 41:44, 5))|
                                      
                                      applicant_race.1 %in% c(4, 41:44) 
                                    & (applicant_race.2 %in% c(1, 2, 21:27, 3, 5)|
                                         applicant_race.3 %in% c(1, 2, 21:27, 3, 5)|
                                         applicant_race.4 %in% c(1, 2, 21:27, 3, 5)|
                                         applicant_race.5 %in% c(1, 2, 21:27, 3, 5))|
                                      
                                      applicant_race.1 %in% c(5) 
                                    & (applicant_race.2 %in% c(1, 2, 21:27, 3, 4, 41:44)|
                                         applicant_race.3 %in% c(1, 2, 21:27, 3, 4, 41:44)|
                                         applicant_race.4 %in% c(1, 2, 21:27, 3, 4, 41:44)|
                                         applicant_race.5 %in% c(1, 2, 21:27, 3, 4, 41:44))  
                                    ,6,
         if_else(applicant_race.1 == 1 & (is.na(applicant_ethnicity.2)|is.na(applicant_ethnicity.3)|is.na(applicant_ethnicity.4)|is.na(applicant_ethnicity.5)), 1, #aian
               if_else(applicant_race.1 %in% c(2,21:27) 
               & (is.na(applicant_ethnicity.2)|is.na(applicant_ethnicity.3)|is.na(applicant_ethnicity.4)|is.na(applicant_ethnicity.5)), 2, #asian
                
               if_else(applicant_race.1 == 3 
                & (is.na(applicant_ethnicity.2)|is.na(applicant_ethnicity.3)|is.na(applicant_ethnicity.4)|is.na(applicant_ethnicity.5)), 3, #black

                if_else(applicant_race.1 %in% c(4, 41:44) 
                & (is.na(applicant_ethnicity.2)|is.na(applicant_ethnicity.3)|is.na(applicant_ethnicity.4)|is.na(applicant_ethnicity.5)), 4, #nhopi

                if_else(applicant_race.1 == 5  
                & (is.na(applicant_ethnicity.2)|is.na(applicant_ethnicity.3)|is.na(applicant_ethnicity.4)|is.na(applicant_ethnicity.5)), 5, NULL)))))))) #white
                

```


```{r Recoding Variables, include = FALSE}

#Re-coding Ethnicity to 1/0####

data3 = 
  data2 %>%
    mutate(hispanicLatino = if_else(applicant_ethnicity.1 == 2, 0,
                                     if_else(is.na(applicant_ethnicity.1),2,1)),
           aian = if_else(applicant_race.1 == 1 & (is.na(applicant_ethnicity.2)|is.na(applicant_ethnicity.3)|is.na(applicant_ethnicity.4)|is.na(applicant_ethnicity.5)), 1, 0),
         
        asian = if_else(applicant_race.1 %in% c(2,21:27) & 
(is.na(applicant_ethnicity.2)|is.na(applicant_ethnicity.3)|is.na(applicant_ethnicity.4)|is.na(applicant_ethnicity.5)), 1, 0),

        black = if_else(applicant_race.1 == 3 & 
(is.na(applicant_ethnicity.2)|is.na(applicant_ethnicity.3)|is.na(applicant_ethnicity.4)|is.na(applicant_ethnicity.5)), 1, 0),

        nhopi = if_else(applicant_race.1 %in% c(4, 41:44) & 
(is.na(applicant_ethnicity.2)|is.na(applicant_ethnicity.3)|is.na(applicant_ethnicity.4)|is.na(applicant_ethnicity.5)), 1, 0),

        white =  if_else(applicant_race.1 == 5 & 
(is.na(applicant_ethnicity.2)|is.na(applicant_ethnicity.3)|is.na(applicant_ethnicity.4)|is.na(applicant_ethnicity.5)), 1, 0),

      multiRace = if_else(applicant_race.1 == 1 
                          & (applicant_race.2 %in% c(2,21:27, 3, 4, 41:44, 5) |
                             applicant_race.3 %in% c(2,21:27, 3, 4, 41:44, 5) |
                             applicant_race.4 %in% c(2,21:27, 3, 4, 41:44, 5) |
                             applicant_race.5 %in% c(2,21:27, 3, 4, 41:44, 5))|



                             applicant_race.1 %in% c(2, 21:27) 
                          & (applicant_race.2 %in% c(1, 3, 4, 41:44, 5)|
                             applicant_race.3 %in% c(1, 3, 4, 41:44, 5)|
                             applicant_race.4 %in% c(1, 3, 4, 41:44, 5)|
                             applicant_race.5 %in% c(1, 3, 4, 41:44, 5))|
                            
                             applicant_race.1 %in% c(3) 
                          & (applicant_race.2 %in% c(1, 2, 21:27, 4, 41:44, 5)|
                             applicant_race.3 %in% c(1, 2, 21:27, 4, 41:44, 5)|
                             applicant_race.4 %in% c(1, 2, 21:27, 4, 41:44, 5)|
                             applicant_race.5 %in% c(1, 2, 21:27, 4, 41:44, 5))|
                               
                             applicant_race.1 %in% c(4, 41:44) 
                          & (applicant_race.2 %in% c(1, 2, 21:27, 3, 5)|
                             applicant_race.3 %in% c(1, 2, 21:27, 3, 5)|
                             applicant_race.4 %in% c(1, 2, 21:27, 3, 5)|
                             applicant_race.5 %in% c(1, 2, 21:27, 3, 5))|
                            
                             applicant_race.1 %in% c(4) 
                          & (applicant_race.2 %in% c(1, 2, 21:27, 3, 4, 41:44)|
                             applicant_race.3 %in% c(1, 2, 21:27, 3, 4, 41:44)|
                             applicant_race.4 %in% c(1, 2, 21:27, 3, 4, 41:44)|
                             applicant_race.5 %in% c(1, 2, 21:27, 3, 4, 41:44))  
                             ,1, 0)) 
  
  

```

For the rest of the analysis, I will be using a new variable, `race``. The applicant race were recoded into:
1 = American Indians/Alaska Natives
2 = Asian,
3 = Black, 
4 = Native Hawaiians and Other Pacific Islanders, 
5 = White, and 
6 = MultiRace. 

Ethnicity was recoded into the 0/1 hispanicLatino variable.

Looking at 2 cross-tabs to get an expected percentage of a fair breakdown of actions taken:
  1) Action taken by state
  2) Action taken by state and recoded race categories


```{r Actions Taken by State (Counts and Percentage), include = FALSE}

actionTakenstats = 
  data3 %>%
    group_by(action_taken) %>%
      summarize(count = n()) %>%
        mutate(actionTakenpct = scales::percent(count/sum(count), accuracy = 0.1))

fileClosedstats = 
  actionTakenstats %>%
    filter(action_taken == 5)

purchasedStats = 
    actionTakenstats %>%
    filter(action_taken == 6)

actiontakeStats =   
  data3 %>%
      group_by(state_code, action_taken) %>%
        summarise(countVar = n()) %>%
          mutate(pctVar = scales::percent(countVar/sum(countVar), 0.1),
                 countVar = scales::comma(countVar))

```
Across the 2 states and DC, the majority of loans in the data set were purchased (`r purchasedStats$actionTakenpct`). The variance in the percentages across the 3 jurisdictions is small, with DC having the lowest percentage of purchased loans at 80.9% and Virginia with the highest at 81.3%. Next, I will look at actions take by state and by race. 

```{r Printing the Actions Taken by State Statistics, echo = FALSE}

kable(actiontakeStats, col.names =  c("State", "Action Taken", "N", "%"), align = "ccrr")

```


```{r Number of Missing Race Entries, include = FALSE}

missingRacecount = 
  data3 %>%
    filter(is.na(race)) %>%
      group_by(race) %>%
        summarize(N = n()) %>%
        mutate(N = scales::comma(N))

missingRacepct =
 scales::percent(nrow(data3 %>%
    filter(is.na(race)))/nrow(data3), 0.1)

```

There are a large number of missing race data (`r missingRacecount$N`) which will limit the insights for this analysis. In order to get a more accurate breakdown of the action taken percentage, I will remove observations with missing race information.

```{r Data Set Without Missing Race Data}

raceDataset = 
  data3 %>%
    filter(!is.na(race))

```


```{r Action Taken without Missing Race Data, warning = FALSE}

purchasedStatsrace =
  raceDataset %>%
    count(action_taken) %>%
      mutate(pct = scales::percent(n/sum(n), 0.1)) %>%
        filter(action_taken == 6)

kable(raceDataset %>%
  group_by(state_code, action_taken) %>%
    summarize(count = n()) %>%
        mutate(actionTakenpct = scales::percent(count/sum(count), accuracy = 0.1)), 
  col.names = c("State", "Action Taken", "N", "%"))

```

After removing the observations with missing race data, the action taken percent decreases from `r purchasedStats$actionTakenpct` to `r purchasedStatsrace$pct`, with DC having the highest purchased percentage at 65.2% and Maryland with the lowest percentage at 53.7%.


```{r Race, State, Actions Taken Statistics}

kable(raceDataset %>%
      group_by(state_code, race, action_taken) %>%
       summarize(countVar = n()) %>%
          mutate(pctVar = scales::percent(countVar/sum(countVar), 0.1),
                 countVar = scales::comma(countVar)), col.names = c("State", "Race", 
                                                                    "Action Taken", "N", "%"), 
      align = "ccccc")
```

From this table, we can see variability in the percentages of action taken by race. Please note that there are small cell sizes. Most race groups had higher percentages of 5 - File closed for incompleteness, with the exception of White applicants in all jurisdictions.  Multirace and Asian applicants had higher purchase percentages  in DC.

```{r Statistically Testing the Difference in Actions Taken by State, include = FALSE}


stateChisq = chisq.test(raceDataset$state_code, raceDataset$action_taken)


```

The chi square test for action taken by state was statistically significant, with a p-value of `r stateChisq$p.value`, meaning that there are differences in the observed versus expected counts  in action taken outcomes by jurisdiction.

# Limitations & Next Steps

Due to a large number of missing race entries (`r missingRacecount$N` or 
`r missingRacepct` of observations), we do not have a full data set at our disposal to conduct analyses by race. Similarly, there were race groups - American Indian/Alaska Natives in particular - with single-digit cells by state. To overcome these limitations, more years and states (especially Alaska and southwestern states with higher AI/AN population) can be added to this work.

Including observations with other loan outcomes (i.e., 1-4, 7, & 8) will give a more complete picture of loan disposition and not only incomplete or purchased.

After adding more data, this program can be used to re-run current analyses and to continue on with a logistic regression in order to further understand the relationship between race and loan outcomes.
